Title Text

### 2013-05-17 papers/gpce2013/gpce2013.tex
Figures
we

\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintainance effort: once you decided to implement a feature on
client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
Even worse, logic parts that need to run on both client-side and server-side are duplicated. For instance, HTML
fragments may be built from the server-side when a page is requested by a client, but they may also be built from the
client-side to perform an incremental update subsequent to an user action. How could developers write HTML fragment
definitions once and render them on both client-side and server-side? The more interactive the application is, the
more logic needs to be duplicated between the server-side and the client-side (explain why?).

@@

@@

@@

@@

@@

We propose to achieve \emph{ad hoc} polymorphism using
typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
retroactive extension without modifying objects prototypes because it is type-directed: the dispatch happens at
compile-time rather than at runtime.

// Polymorphic function
def listWidget[A : Show](items: Rep[List[A]]): Rep[Node] = {
  tag("ul")(
    for (item <- items) yield {
      tag("li")(implicitly[Show[A]].show(item))
    }
  )
}

// Type `User`
type User = Record { val name: String; val age: Int }
// Implementation of Show for a User
implicit val showUser = Show[User] { user =>
  tag("span", "class"->"user")(
    text(user.name + "(" + user.age + " years)")
  )
}

// Type `Article`
type Article = Record { val name: String; val price: Double }
// Implementation of Show for an Article
implicit val showArticle = Show[Article] { article =>
  tag("span")(
    text(article.name),
    tag("strong")(text(article.price + " Euros"))
  )
}

// Main program
def main(users: Rep[List[User]], articles: Rep[List[Article]]) = {
  document.body.append(listWidget(users))
  document.body.append(listWidget(articles))
}
\end{lstlisting}
\end{figure}

@@

The previous work on js-scala \cite{Kossakowski12_JsDESL} showed how to make asynchronous programming more
convenient by making asynchronous calls looking like synchronous calls (\ie{} returning a value instead of taking a
callback as parameter). This work was based on the Scala continuations compiler plugin. We claim that, although this
programming model removes the “callback hell”, it can make the code hard to reason about because there is no explicit
distinction between synchronous and asynchronous computations.

We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
turning them into first-class citizens. The DSL is monadic so we can solve the callback hell problem thanks to the
Scala \code{for} notation.

For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written using our DSL.
The \code{createVertex} function now returns an asynchronous value instead of taking a callback as
parameter. Then, the \code{for} expression allows us to get the vertex, when available, and to insert it on the user
interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
a parameter, the code is easier to modularize into loosely coupled parts.

\begin{figure}
\centering
\begin{lstlisting}[caption=Asynchronous values are first class citizen,label=async-first-class]
def createVertex(text: Rep[String]): Rep[Future[Vertex]] =
  Ajax.post[Vertex]("/create", new Record { val content = text })

for (vertex <- createVertex("Hello, World!")) {
  addVertex(vertex)
}
\end{lstlisting}
\end{figure}

\begin{figure}
\centering
\begin{lstlisting}[caption=No callback hell,label=async-no-callback]
for {
  foo <- Ajax.get(fooUrl())
  bar <- Ajax.get(barUrl(foo))
  _ <- future(println("bar = " + bar))
  baz <- Ajax.get(bazUrl(bar))
} println(foo + bar + baz)
\end{lstlisting}
\end{figure}

Listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
first Ajax request is available, the next statement is executed, and so on. There is no nested callbacks and the
order of execution is directly reflected by the order of statements.

An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
\begin{itemize}
\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} also eventually transforms the value when
available~;
\end{itemize}

Asynchronous values can be used and transformed using the \code{foreach}, \code{map} and \code{flatMap} methods,
turning them into first-class
citizens: functions can take them as parameters and return them. Consequently, it is easier to write concurrent code:
listing \ref{async-parallel-1} shows how to run two parallel asynchronous computations and to do something with their
results when both are available. Writing equivalent code in pure JavaScript is tedious because it requires that the
callback passed to each asynchronous computation checks if the other has been completed before, and to store the
value, when availalbe, in a shared place (so both callbacks can get the other’s value).

\begin{figure}
\begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
val fooAsync = Ajax.get("/foo")
val barAsync = Ajax.get("/bar")
for {
  foo <- fooAsync
  bar <- barAsync
} println(foo + bar)
\end{lstlisting}
\end{figure}

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider the following typical JavaScript code finding a particular widget in the page and a then particular
button in the widget:

\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget = document.querySelector("div.login");
var loginButton = loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", function (e) { ... });
\end{lstlisting}

@@

\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
var loginWidget = document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton = loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.addEventListener("click", function (e) { ... });
  }
}
\end{lstlisting}

@@

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
function to define a tag and a \code{text} function to define a text node.

\begin{figure}
\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
def vertexDom(v: Rep[Vertex]) =
    tag("g", "class"->"vertex",
             "transform"->("translate("+v.x+","+v.y+")"))(
        tag("rect", "width"->v.width, "height"->v.height)(),
        tag("text", "width"->v.width, "height"->v.height)(
            text(v.content)
        )
    )
\end{lstlisting}
\end{figure}

@@

@@

@@

>>>>

\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{graphicx}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintainance effort: once you decided to implement a feature on
client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
Even worse, logic parts that need to run on both client-side and server-side have to be duplicated. For instance,
HTML fragments may be built from the server-side when a page is requested by a client, but they may also be built
from the client-side to perform an incremental update subsequent to an user action. How could developers write HTML
fragment definitions once and render them on both client-side and server-side? The more interactive the application
is, the more logic needs to be duplicated between the server-side and the client-side (explain why?).

@@

Developers express programs using a language that is translated into a form executable on a target platform. This
section presents different approaches for defining cross-platform languages.

@@

One way consists in hard-coding, in the language’s compiler, the transformation to each target platform. Figure
\ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because
a lot of concepts are defined at the language level. Examples of such languages are Links~\cite{Cooper07_Links},
Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the
compiler, and modifying a compiler requires a high effort. Furthermore, these languages also require to support
common abstraction and composition mechanisms, as general purpose languages do, in order to make it easier to write
real programs. So they need to be at least general purpose languages. It means that for every problem you have to
rewrite a full-featured programming language before addressing the concepts specific to the problem domain. This
approach does not scale.

\begin{figure}
  \centering
  \includegraphics[width=3cm]{fat.pdf}
  \caption{Fat language}
  \label{fat-lang}
\end{figure}

\subsection{Domain Specific Languages}

Problem: interoperability.

@@

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language. Figure \ref{thin-lang} depicts this approach. The general purpose language is used as a host language that
does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
this approach gives no opportunity to translate a concept efficiently, according to the host platform
characteristics.

\begin{figure}
  \centering
  \includegraphics[width=3cm]{thin.pdf}
  \caption{Thin language}
  \label{thin-lang}
\end{figure}

@@

Figure \ref{dedsl} shows a last approach that tries to get the benifits of the two previous approaches, without
their shortcomings.

\begin{figure}
  \centering
  \includegraphics[width=3cm]{lms.pdf}
  \caption{Deeply embedded languages}
  \label{dedsl}
\end{figure}

@@

@@

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider the following typical JavaScript code finding a particular widget in the page and then a particular
button in the widget:

\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget =
  document.querySelector("div.login");
var loginButton =
  loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", handler);
\end{lstlisting}

@@

\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
var loginWidget =
  document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton =
    loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.
      addEventListener("click", handler);
  }
}
\end{lstlisting}

@@

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{el}
function to define a tag and any \code{String} value is considered to be a text node.

\begin{figure}
\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
def vertexDom(v: Rep[Vertex]) =
    el("g", "class"->"vertex",
            "transform"->("translate("+v.x+","+v.y+")"))(
        el("rect", "width"->v.width, "height"->v.height)(),
        el("text", "width"->v.width, "height"->v.height)(
            v.content
        )
    )
\end{lstlisting}
\end{figure}

@@

We propose to achieve \emph{ad hoc} polymorphism using
typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
compiler, so there is no runtime overhead.

// Polymorphic function
def listWidget[A](items: Rep[List[A]])
      (implicit A: Show[A]): Rep[Node] =
  el("ul")(
    for (item <- items) yield {
      el("li")(A.show(item))
    }
  )

// Type `User`
type User = Record {
  val name: String
  val age: Int
}
// Implementation of Show for a User
implicit val showUser = Show[User] { user =>
  el("span", "class"->"user")(
    user.name + "(" + user.age + " years)"
  )
}

// Main program
def main(users: Rep[List[User]]) = {
  document.body.append(listWidget(users))
}
\end{lstlisting}
\end{figure}

@@

See http://github.com/js-scala/js-scala.

@@

We implemented several applications using js-scala. We also have written several implementations of a complete
application, using different approaches to write the client and server sides, and compared the amount of code
written, the runtime performances and the ability to modularize the code and to maintain it.
